---
title: "Puerto Rico Migration Flows"
author: "Wendy Prudencio"
---

# Overview

Puerto Rico is a territorial island belonging to the United States and is home to an estimated 3.2 million people (Census, n.d). The archipelago is located in the Caribbean Sea and is susceptible to hurricanes, landslides, mudslides, earthquakes, and other natural hazards. In the last decade, Puerto Rico has experienced three major hurricanes, Irma, and Maria in 2017 and Fiona in 2022. Hurricanes Irma and Maria, category 4 and 5, respectively, destroyed critical infrastructure, hospitals, and damaged thousands of homes displacing thousands of Puerto Ricans.

The primary goal of this project is research, compare, analyze, and document migration flows from Puerto Rico to the continental U.S. pre-and-post hurricanes Irma and Maria and possibly Fiona, if data is available.

# Data and approach

The data to support the goal of this project that I currently have includes data from the Census Bureau's Population Estimate Program (PEP) from 2015-2019. The Census Bureau releases data on population and housing on a “flow basis” to estimate changes from the decennial (10-year) census to the present year (Census, 2022). The PEP estimates are collected at the country, state, city, and municipality levels for all of the U.S. (Census, 2022). This data is used by the federal government to allocate federal funds throughout the country. 

In addition, Mapbox will be used as a visual tool to show migration flows from Puerto Rico to the continental U.S. Mapbox was designed to customize maps and provide tools for web-based applications (mapbox, n.d.)

My approach is to explore the data and determine how to best analyze the data using what I have found thus far and be able to combine methodology to produce a visual.

## Questions

The questions I intend to answer are: 1. Which year had the most migration from Puerto Rico to the continental U.S.? 2. What continental U.S. state received the most influx of Puerto Rican individuals? 

## Description of data

1.    The first dataset is from the Census Bureau’s with estimated population data for municipalities and census tracts. PEP data is obtained through the U.S. Census Bureau and has variables of moved-in and moved-out of different areas worldwide (Census, n.d.). The PEP data used for this project is from 2015-2019. The dataset will be used to compare the number of individuals living on the island pre-and-post Hurricanes Irma and Maria. The dataset includes the variables, GEOID1 is the Puerto Rico census tract code, GEOID2 is a geographic code, location, estimates, and margin of error. 

I do not think there are any ethical issues around the collection of any of the datasets and I do not anticipate any ethical issues from the work I plan on doing with the data. 

## Preliminary Analysis

I started my exploratory analysis by using the R packages tidycensus, tidyverse, and tigris in RStudio to use PEP data to explore some population estimates and characteristics such as age and gender. Using the walkerke/tidycensus vignette from Github, I was able to start the exploratory migration flows in Puerto Rico using `get_flows()` function that provides movedin, movedout, and movednet variables with a margin of error for each column. The `get_flows()` function only uses the geography as county, county subdivision, or metropolitan statistical area, unfortunately it does not use state. Puerto Rico has 78 municipalities/ counties; therefore, I will start comparing the municipalities in 2015.

## Formal Analysis

A data frame was created using the `purrr::map_dfr()` function with PEP data for every year between 2015 and 2019. The first question: which year had the most migration from Puerto Rico to the continental U.S.? To answer this question, data was filtered to exclude "Africa", "Asia", "Central America", "Caribbean", "Europe", "South America", and "Puerto Rico". These variables were filtered out because they are non-U.S. locations and will skew the results. Puerto Rico had the highest migration flow to the continental U.S. in 2018 with 215,457 people moving out. To answer the second question: what continental U.S. state received the most influx of Puerto Rican individuals? Zavala County, Texas. 

The ‘mapdeck’ function was used to create a visualization of the out flow migration from Puerto Rico to the continental U.S. I used San Juan, the capital of Puerto Rico and Mayagüez, a major city on the west coast of Puerto Rico to visualize the migration patterns. 

## Discussion

Patterns of migration flows are difficult to trace, but with data such as the PEP allows researchers to attempt to estimate when and where did people move. This project was able to provide the foundation to migration flows in Puerto Rico after Hurricanes Irma and Maria in 2017.

# References

Mapbox. (n.d.). Our vision. Retrieved from https://www.mapbox.com/about/company/

U.S. Census Bureau. (n.d.). Quickfacts Puerto Rico. Retrieved from https://www.census.gov/quickfacts/PR

U. S. Census Bureau. (2022). Population and Housing Unit Estimates Tables. Retrieved from https://www.census.gov/programs-surveys/popest.html

# Links

PEP data is provided by: (https://www.census.gov/programs-surveys/popest.html)

NFIP data is provided by: (https://www.fema.gov/about/openfema/data-sets#nfip)

Mapbox: (https://www.mapbox.com)

```{r setup}
#| echo: false
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
library(readxl)

options(tigris_use_cache = TRUE)
```

```{r}
library(usethis)
```

```{r}
usethis::use_git_remote("origin", url = NULL, overwrite = TRUE)
git_sitrep()
```


git commit --no verify

```{r}
use_github(organisation = "bldgspatialdata22")
```

```{r}
use_readme_rmd()

use_cc0_license()
```

```{r}
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
census_api_key("3efcfb0ddc4c895edaac8221c73a40ca59dd392c")
"install = TRUE"
```

## Census population estimates for sex, age, and latino/a/x orgin in Puerto Rico

```{r}
pr_age_hisp <-
  get_estimates(
    geography = "state",
    product = "characteristics",
    breakdown = c("SEX", "AGEGROUP", "HISP"),
    breakdown_labels = TRUE,
    state = "PR"
  )
```

```{r}
pr_age_hisp
```

## Migration flow per county/ municpality for 2015-2019

````{r}
library(readr)

read_rds("pr_counties_flows_2015_2019.rds", refhook = NULL)
```

```{r}
pr_movedout_summary <-
pr_counties_flows_2015_2019 %>%
  filter(!(FULL2_NAME %in% c("Africa", "Asia", "Central America", 
  "Caribbean", "Europe", "South America", "Puerto Rico" )), variable == "MOVEDOUT") %>% 
  group_by(year) %>%
  summarise(
    total_movedout = sum(estimate, na.rm = TRUE)
  )
pr_movedout_summary

pr_movedout_summary %>% 
  ggplot() +
  geom_col(aes(as.factor(year), total_movedout)) +
  guides(fill = "none") +
  xlab("Year") +
  ylab("Population Totals") +
  ggtitle("Total Migration Flow in Puerto Rico by Year") +
  ggeasy::easy_center_title()

ggplot(pr_movedout_summary, aes(x=year, y=total_movedout)) + geom_point() +
  xlab("Year") +
  ylab("Population Totals") +
  ggtitle("Total Migration Flow in Puerto Rico by Year in Points") +
  ggeasy::easy_center_title()
```

```{r}
pr_movedout_state <-
  pr_counties_flows_2015_2019 %>%
  filter(!(FULL2_NAME %in% c("Africa", "Asia", "Central America", 
  "Caribbean", "Europe", "South America", "Puerto Rico",variable == "MOVEOUT"))) %>%
  group_by(FULL2_NAME) %>%
  summarise(
    total_movedout = sum(estimate, na.rm = TRUE))

pr_movedout_state

max(pr_movedout_state$FULL2_NAME)

max(pr_movedout_state$total_movedout)

pr_movedout_state %>% 
  ggplot() +
  geom_col(aes(as.factor(FULL2_NAME), total_movedout)) +
  theme_grey() +
  guides(fill = "none") +
  xlab("Counties") +
  ylab("Population Totals") +
  ggtitle("Total Migration Flow from Puerto Rico to Other Places") +
  ggeasy::easy_center_title()
```

```{r}
library(mapdeck)

sanjuan_inflow <-
  get_flows(
  geography = "county",
  state = "PR",
  county = "San Juan Municipio",
  geometry = TRUE
) %>%
  filter(variable == "MOVEDIN") %>%
  na.omit() %>%
  arrange(desc(estimate))

mapboxapi::mb_access_token("pk.eyJ1Ijoid3BydWRlbmNpbyIsImEiOiJjbGJibWoxYmcxZnV3M29tcmVnbzMzbzMxIn0.4O0UofCM_KbVsb_EZM-i6g", overwrite=TRUE)

sanjuan_inflow %>%
  slice_max(estimate, n = 30) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = mapboxapi::get_mb_access_token("pk.eyJ1Ijoid3BydWRlbmNpbyIsImEiOiJjbGJibWoxYmcxZnV3M29tcmVnbzMzbzMxIn0.4O0UofCM_KbVsb_EZM-i6g")) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight",
          update_view = FALSE) 

mayaguez_inflow <-
  get_flows(
  geography = "county",
  state = "PR",
  county = "Mayaguez Municipio",
  geometry = TRUE
) %>%
  filter(variable == "MOVEDIN") %>%
  na.omit() %>%
  arrange(desc(estimate))


mayaguez_inflow %>%
  slice_max(estimate, n = 30) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = mapboxapi::get_mb_access_token("pk.eyJ1Ijoid3BydWRlbmNpbyIsImEiOiJjbGJibWoxYmcxZnV3M29tcmVnbzMzbzMxIn0.4O0UofCM_KbVsb_EZM-i6g")) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight",
          update_view = FALSE) 

sanjuan_mayaguez_inflow <-
 bind_rows(
   sanjuan_inflow,
   mayaguez_inflow,
 )

sanjuan_mayaguez_inflow %>%
  slice_max(estimate, n = 30) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = mapboxapi::get_mb_access_token("pk.eyJ1Ijoid3BydWRlbmNpbyIsImEiOiJjbGJibWoxYmcxZnV3M29tcmVnbzMzbzMxIn0.4O0UofCM_KbVsb_EZM-i6g")) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight",
          update_view = FALSE) 
```
